---
title: "Probability Distributions"
author: "David Apamo"
date: "`r Sys.Date()`"
output: html_document
---

# 1.Bernoulli distribution {X ∼ BERN(p)}

The Bernoulli distribution is the probability distribution of a discrete random variable that has only two possible outcomes, i.e such as success or failure.

```{r}
# The probability of success(getting a head) in one toss of a fair coin
dbinom(x = 1, size = 1, prob = 0.5)
# The probability of getting a tail
dbinom(x = 0, size = 1, prob = 0.5)
# Note that size(n) = 1 for a Bernoulli trial
```
```{r}
# The probability of getting a tail in one toss of a bias coin, with p = 0.3
dbinom(x = 0, size = 1, prob = 0.3)
```

# 2.Binomial distribution {X ∼ BIN(n, p)}; mean = np, var = np(1-p)

The binomial distribution is the distribution of successes in n number of trials
involving binary discrete random variables.

```{r}
## Rolling a fair die example

# What is the probability of getting a 4 when you roll a die 8 times?
dbinom(x = 4, size = 8, prob = (1/6))
# The probability of face 5 showing up
dbinom(x = 5, size = 8, prob = (1/6))
```
```{r}
# Full PMF for the die 
Die_pmf <- dbinom(x = 0:8, size = 8, prob = (1/6))
Die_pmf

# Confirm if sum adds up-to 1
sum(Die_pmf)
# Round off the probabilities to 4sf
round(Die_pmf,4)
```
```{r}
# Mean and variance
np <- (8*(1/6))
np
npq <- (8*(1/6)*(5/6))
npq
```
```{r}
# Plot the probability mass function
barplot(Die_pmf, names.arg = 0:8, space = 0, xlab = "x", ylab = "Pr(X = x)")
# Face 1 has the highest probability of showing up (more than 0.35)
```
```{r}
# Pr(X <= 3)

# First alternative
sum(dbinom(x = 0:3,size = 8, prob = (1/6)))
# Second alternative
pbinom(q = 3, size = 8, prob = (1/6))

# Pr(X >= 3)
1 - pbinom(q = 3, size = 8, prob = (1/6))
```
```{r}
# Finding q, given p = 0.95
qbinom(p = 0.95, size = 8,prob = 1/6)

# Generating random numbers from a binomial distribution
rbinom(n = 3, size = 8, prob = (1/6))
```

# 3.Poisson distribution {X ∼ POIS(λp)}; mean = λ, var = λ

Poisson distribution is a discrete probability distribution which gives the probability of an event happening a certain number of times within a given time interval. The values of X can only take whole values. This distribution has only one parameter λ.

```{r}
# Given λ = 3.22, find Pr(X = 3)
dpois(x = 3,lambda = 3.22)
# Pr(x = 7)
dpois(x = 7, lambda = 3.22)
# Pr(0 <= X <= 10), to 4sf
round(dpois(0:10,3.22),4)
```
```{r}
# Visualize the PMF Pr(0 <= X <= 10)
barplot(dpois(x = 0:10, lambda = 3.22), ylim = c(0,0.25), space = 0, 
        names.arg = 0:10, ylab = "Pr(X=x)", xlab = "x", 
        main = "Probability mass function")

```
```{r}
# Cumulative Mass Function

# Pr(X ≤ 2)
ppois(q = 2,lambda = 3.22)

# Pr(X >= 6)
1 - ppois(q = 5,lambda = 3.22)

# Visualize the Cumulative mass function
barplot(ppois(q = 0:10,lambda = 3.22), ylim = 0:1, space = 0, 
        names.arg = 0:10, ylab = "Pr(X<=x)", xlab="x", 
        main = "Cumulative Mass Function")

```
```{r}
# Generate random numbers from a poisson distribution
rpois(n = 15,lambda = 3.22)
```

# 4.Uniform distibution {X ∼ UNIF(a, b)}; mean = ((a+b)/2), var = ((b-a)^2/12)

```{r}
# X ∼ UNIF(−0.4,1.1)
# Note that dunif() function produces the value for the function f(x) and not really probability.
dunif(x = c(-2,-0.33,0,0.5,1.05,1.2), min = -0.4, max = 1.1)
# Note that the values that fall outside the bounds defined by min and max are evaluated to zero

# X ∼ UNIF(0.223,0.410) 
dunif(x = c(0.3,0,0.41), min = 0.223, max = 0.41)

```
```{r}
# CMF
# Pr(X <= -0.21)
punif(-0.21, min = -0.4, max = 1.1)
# Pr(X > 0.6)
1 - punif(q = 0.6,min = -0.4, max = 1.1)
```
```{r}
# Generate random numbers from a uniform distribution
runif(n = 10, min = -0.4, max = 1.1)

```

# 5.Normal Distribution {X ∼ N(µ,σ)}

```{r}
# Create values of X
xvals <- seq(-2,2, length = 10 )
xvals
fx <- dnorm(xvals) #Produces a Standard Normal Function by default
fx

# f(x1) ~ N(0,0.64)
fx1 <- dnorm(xvals, mean = 0, sd = 0.8)
fx1
```
```{r}
# Plot the distribution
pacman::p_load(ggplot2)

# create a data frame of X and F(x) values
stdnorm <- data.frame(xvals,fx)
# plot
ggplot(stdnorm, aes(x = xvals, y = fx)) + geom_point() + 
  ggtitle("Standard Normal Distribution")

# create a data frame of X and F(x1) values
fnormal <- data.frame(xvals,fx1)
# plot
ggplot(fnormal, aes(x = xvals, y = fx1)) + geom_point() + 
  ggtitle("Normal Distribution")

```

Standard normal distribution has a broad tail.

```{r}
# Find the probability that lies within 1 standard deviation 
pnorm(q = 1) - pnorm(q = -1)

# The probability lying outside the middle region
(1 - (pnorm(q = 1) - pnorm(q = -1)))

# Given X ~ N(-3.42,0.2), find the probability that (X > -3.22)
1 - pnorm(-3.22, -3.42, 0.2)

# Find the probability that (X < -3.62)
pnorm(-3.62, -3.42, 0.2)
```
```{r}
# Find the quantile value that will give a lower-tail probability of 0.159
qnorm(p = 0.159, mean = -3.42, sd = 0.2) #gives X to be approximately -3.62
# Find the upper quartile (the value above which you'd find a probability of 0.25)
qnorm(p = (1 - 0.25), mean = -3.42, sd = 0.2)
```
```{r}
# Generating random numbers from a normal distribution; X ~ N(-3.42,0.2)
rnorm(n = 7, mean = -3.42, sd = 0.2)
# Generating random numbers from a standard normal distribution
rnorm(n = 7)
```
```{r}
# Generate random numbers from a standard normal distribution and use the qqplot() function to assess the degree of normality
data1 <- rnorm(n = 71)
data2 <- rnorm(n = 710)

qqnorm(data1, main = "Normal Q-Q plot of generated N(0,1) data; n = 71")
qqline(data1,col="gray")
qqnorm(data2,main = "Normal Q-Q plot of generated N(0,1) data; n = 710")
qqline(data2,col="gray")
```

The data with n = 710 produces a lesser random variation. Both plots however, have discrepancies in the tails.


# 6.Exponential Distribution; {X ∼ EXP(λe)}, f(x) = λe exp{−λe x}; 0 ≤ x < ∞

```{r}
# Plot an exponential density function with λ = 1.65
yvals <- seq(0,10,length = 100) # create values for Y

# plot the pdf
plot(yvals, dexp(x = yvals, rate = 1.65), xlim=c(0,10), ylim = c(0,1.65), 
     type="l", xlab = "x", ylab = "f(x)", 
     main = "The distribution of Exponential density function with λ = 1.65")
```
```{r}
# Given the average number of cars passing an individual within a 120-minute window was said to be 107. Define the random variable X to be the waiting time between two consecutive cars passing and, using an exponential distribution for X on a minute time scale, set λ = (107/120)

# Find the probability that the individual will wait for 2.5 minutes before a car passes
# Pr(X > 2.5)
1 - pexp(q = 2.5,rate = (107/120))

# Find the probability that the individual will wait for less than 25 seconds before a car passes
# Find Pr(X < 25 seconds)
pexp(q = (25/60), rate = (107/120))

# Find quantile function/ the cutoff point for the shortest 15 percent of waits.
qexp(p = 0.15, rate = (107/120)) #roughly 0.182 minutes

# The cutoff point for the shortest 25 percent of waits
qexp(p = 0.25, rate = (107/120)) #lower quantile function

```

There's only 10% chance that one waits more than two and a half minutes before another car passes

```{r}
# Generate random numbers from an exponential distribution
rexp(n = 10, rate = (107/120)) # λ = (107/120)
rexp(n = 15, rate = 0.5) # λ = 0.5
```
